---
title: Advanced: Data Serialization and JavaScript Injection
---



```{r, include = FALSE}
litedown::reactor(
  collapse = TRUE,
  comment = "#> "
)
RUN_EXAMPLES <- nzchar(Sys.getenv("R_ALPINEJS_RUN_EXAMPLES"))
```

```{r setup, eval=TRUE, echo=TRUE}
library(alpinejs)
library(htmltools)
library(jsonlite)
```

```{r test_serialization, eval=RUN_EXAMPLES, echo=FALSE}
# Verify data serialization works
cat("Testing R-to-JSON conversion...\n")
# Test basic types
test_data <- list(
  count = 42,
  ratio = 3.14,
  active = TRUE,
  name = "Alice",
  items = c(1, 2, 3)
)
json_str <- jsonlite::toJSON(test_data)
cat("JSON conversion successful\n")
cat("Sample output (first 100 chars):", substr(as.character(json_str), 1, 100), "\n")
```

# Advanced: Data Serialization and JavaScript Injection

This vignette covers how alpinejs converts R data types to JSON and how to inject raw JavaScript expressions.

## Type Conversion Rules

When you call `alpine_data()`, alpinejs serializes R objects to JSON using specific rules:

### Atomic Types

```{r}
# Numeric, logical, character -> JSON equivalents
result <- alpine_tag("div") |>
  alpine_data(
    count = 42,            # number -> 42
    ratio = 3.14,          # numeric -> 3.14
    active = TRUE,         # logical -> true
    name = "Alice"         # character -> "Alice"
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

### Vectors

Unnamed vectors become JSON arrays:

```{r}
result <- alpine_tag("div") |>
  alpine_data(
    ids = c(1, 2, 3, 4, 5),                    # -> [1,2,3,4,5]
    names = c("Alice", "Bob", "Charlie"),      # -> ["Alice","Bob","Charlie"]
    enabled = c(TRUE, FALSE, TRUE)             # -> [true,false,true]
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

### Lists

Named lists become JSON objects:

```{r}
result <- alpine_tag("div") |>
  alpine_data(
    user = list(
      name = "Alice",
      age = 30,
      email = "alice@example.com"
    ),
    config = list(
      apiUrl = "https://api.example.com",
      timeout = 5000,
      debug = FALSE
    )
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

This renders as:

```javascript
{
  user: {name: "Alice", age: 30, email: "alice@example.com"},
  config: {apiUrl: "https://api.example.com", timeout: 5000, debug: false}
}
```

### Data Frames

Data frames become arrays of objects (row-oriented to column-oriented):

```{r}
# Simple data frame
df <- data.frame(
  id = 1:3,
  name = c("Alice", "Bob", "Charlie"),
  score = c(95, 87, 92),
  stringsAsFactors = FALSE
)

result <- alpine_tag("div") |>
  alpine_data(data = df)

cat("```html\n", as.character(result), "\n```", sep="")
```

This becomes:

```javascript
{
  data: [
    {id: 1, name: "Alice", score: 95},
    {id: 2, name: "Bob", score: 87},
    {id: 3, name: "Charlie", score: 92}
  ]
}
```

### Special R Types

**Factors** are converted to character vectors:

```{r}
# Factor -> character vector
factors_data <- factor(c("A", "B", "B", "C"))

result <- alpine_tag("div") |>
  alpine_data(categories = factors_data) |>  # -> ["A","B","B","C"]
  htmltools::browsable()

cat("```html\n", as.character(result), "\n```", sep="")
```

**Dates and POSIXct** are converted to ISO 8601 strings:

```{r}
# Date -> ISO 8601 string
today <- Sys.Date()
now <- Sys.time()

result <- alpine_tag("div") |>
  alpine_data(
    date = today,    # -> "2026-02-21"
    datetime = now   # -> "2026-02-21T10:30:45Z"
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

**NA values** become JSON `null`:

```{r}
# NA -> null
result <- alpine_tag("div") |>
  alpine_data(
    value = NA,
    text = NA_character_,
    date = as.Date(NA)
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

Renders as:

```javascript
{value: null, text: null, date: null}
```

## Nested and Complex Structures

You can combine different types by nesting:

```{r}
# Complex nested structure
result <- alpine_tag("div") |>
  alpine_data(
    analytics = list(
      users = data.frame(
        id = 1:3,
        name = c("Alice", "Bob", "Charlie"),
        active = c(TRUE, FALSE, TRUE),
        stringsAsFactors = FALSE
      ),
      summary = list(
        total = 3,
        active = 2,
        lastUpdated = Sys.time()
      ),
      categories = c("A", "B", "B", "C")
    )
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

## Edge Cases and Gotchas

### Empty Vectors and Lists

```{r}
result <- alpine_tag("div") |>
  alpine_data(
    empty_vector = c(),           # -> []
    empty_list = list(),          # -> {}
    null_value = NULL             # -> null
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

### Mixed Types in Vectors

R coerces mixed types; this is preserved in JSON:

```{r}
# Vector with mixed types (coerced to character)
result <- alpine_tag("div") |>
  alpine_data(
    mixed = c(1, "two", 3.5)  # -> ["1","two","3.5"]
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

### Performance with Large Data

For very large data frames (1000+ rows), consider:

1. **Server-side filtering**: Pass only visible rows
2. **Pagination**: Load data on demand
3. **htmlwidgets**: For specialized views (DT tables, Plotly)

```{r}
# Good: subset to first 100 rows
large_df <- data.frame(x = 1:10000, y = rnorm(10000))

result <- alpine_tag("div") |>
  alpine_data(data = head(large_df, 100)) |>  # Only first 100 rows
  htmltools::browsable()

cat("```html\n", as.character(result), "\n```", sep="")
```

## JavaScript Injection with JS()

For custom logic, inject raw JavaScript using `htmlwidgets::JS()`:

```{r}
# Define a method using JS()
result <- alpine_tag("div",
  alpine_tag("button", "Greet",
    alpine_attr(`@click` = "greet()")
  ),
  alpine_tag("p",
    alpine_attr(`x-bind:textContent` = "greeting")
  )
) |>
  alpine_data(
    greeting = "Ready",
    greet = htmlwidgets::JS(
      "function() { this.greeting = 'Hello, Alpine!'; }"
    )
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

### When to Use JS()

Use `JS()` for:

**Functions and Methods**
```{r}
result <- alpine_tag("div") |>
  alpine_data(
    increment = htmlwidgets::JS("function() { this.count++; }"),
    formatDate = htmlwidgets::JS("(date) => new Date(date).toLocaleDateString()")
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

**Arrow Functions**
```{r}
result <- alpine_tag("div") |>
  alpine_data(
    double = htmlwidgets::JS("(x) => x * 2"),
    filter = htmlwidgets::JS("(arr, pred) => arr.filter(pred)")
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

**Complex Expressions**
```{r}
result <- alpine_tag("div") |>
  alpine_data(
    timestamp = htmlwidgets::JS("Date.now()"),
    location = htmlwidgets::JS("window.location.href")
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

**Simple Values** (use plain R instead)
```{r}
# Don't do this (syntactically valid but wasteful):
result <- alpine_tag("div") |>
  alpine_data(name = htmlwidgets::JS('"Alice"')) |>  # Use: name = "Alice"
  htmltools::browsable()

# Better approach:
alpine_tag("div") |>
  alpine_data(count = 5) |>  # Use plain R, not JS()
  htmltools::browsable()

cat("```html\n", as.character(result), "\n```", sep="")
```

## CSP-Safe JavaScript

The default CSP-compliant build of Alpine.js limits what JavaScript expressions you can use. The following expressions are allowed:

### Permitted Expressions (CSP-safe)

```{r}
result <- alpine_tag("div") |>
  alpine_data(
    count = 0,
    doubled = htmlwidgets::JS("this.count * 2"),           # arithmetic
    message = htmlwidgets::JS("this.name + ' is here'"),   # string concat
    result = htmlwidgets::JS("this.x > 10 ? 'Yes' : 'No'") # ternary
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

### Restricted Expressions (CSP-incompatible)

These require `options(alpinejs.use_csp = FALSE)`:

```{r, eval=FALSE}
# NOT CSP-safe: function definition
greet = htmlwidgets::JS("function() { return 'Hi'; }")

# NOT CSP-safe: arrow function
double = htmlwidgets::JS("x => x * 2")

# NOT CSP-safe: constructor
data = htmlwidgets::JS("new Map()")

# NOT CSP-safe: eval/exec
result = htmlwidgets::JS("eval('1+1')")
```

To use unsafe expressions:

```{r}
# Disable CSP mode for the entire session
options(alpinejs.use_csp = FALSE)

# Now arrow functions work
result <- alpine_tag("div") |>
  alpine_data(
    double = htmlwidgets::JS("(x) => x * 2")
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

## Debugging Serialization

To inspect what JSON is generated, use browser dev tools or R itself:

```{r}
# Examine the generated HTML
component <- result <- alpine_tag("div") |>
  alpine_data(
    users = data.frame(name = c("Alice", "Bob"), age = c(30, 25)),
    config = list(debug = TRUE, version = "1.0")
  )

# View raw HTML
as.character(component)  # Shows the x-data attribute value

cat("```html\n", as.character(result), "\n```", sep="")
```

Or check in browser:

```javascript
// Browser console
console.log(Alpine.$data(document.querySelector('div')))  // Inspect component state
```

## Working with Dates and Times

Be explicit about timezone handling:

```{r}
# UTC datetime
utc_time <- as.POSIXct("2026-02-21 12:00:00", tz = "UTC")

# Local time (converted to ISO with Z suffix)
local_time <- Sys.time()

result <- alpine_tag("div") |>
  alpine_data(
    created = utc_time,    # -> "2026-02-21T12:00:00Z"
    modified = local_time   # -> "2026-02-21T10:30:45Z" (or local TZ)
  )

cat("```html\n", as.character(result), "\n```", sep="")
```

In Alpine.js, parse ISO strings back to dates:

```r
alpine_tag("div") |>
  alpine_data(
    created = Sys.time(),
    createdDate = htmlwidgets::JS("new Date(this.created)")
  )
```

## Performance Tips

1. **Avoid huge data frames in alpine_data**: Pass only what's needed for the UI
2. **Use lists for structured data**: More efficient than nested data frames
3. **Minimize nested structures**: Flatten when possible
4. **Cache computed values**: Pre-calculate in R, pass as state

```{r}
# Good: pre-compute statistics
system_status <- list(
  total_users = 1000,
  active_users = 750,
  inactive_users = 250  # Pre-computed instead of calculated in JS
)

# Instead of:
# total_users = 1000,
# active_users = ...,
# inactive_users = htmlwidgets::JS("this.total_users - this.active_users")
```

## Troubleshooting

**Q: "JSON.parse error"**

- Your data contains invalid characters (e.g., backslashes)
- Solution: Let alpinejs handle serialization; don't pre-stringify

**Q: "Undefined variable"**

- Typo in variable name or JS() expression
- Solution: Check browser console; verify spelling in both R and HTML

**Q: Data looks wrong in Alpine**

- Factor converted to character (unexpected)?
- Solution: Convert factors to character explicitly before alpine_data()

**Q: CSP violation in production**

- Using unsafe expressions with CSP mode enabled
- Solution: Set `options(alpinejs.use_csp = FALSE)` or simplify expressions

---

For examples, see the vignettes and `examples/` directory (e.g., `examples/01-todo-list.R`)!
