---
title: Pre-built Components and Patterns
---



```{r, include = FALSE}
litedown::reactor(
  collapse = TRUE,
  comment = "#> "
)
RUN_EXAMPLES <- nzchar(Sys.getenv("R_ALPINEJS_RUN_EXAMPLES"))
```

```{r setup, eval=TRUE, echo=TRUE}
library(alpinejs)
```

```{r test_components, eval=RUN_EXAMPLES, echo=FALSE}
# Verify components exist and return proper structure
cat("Testing built-in components...\n")
acc <- alpine_accordion(items = list(list(title = "Test", content = "Content")))
cat("Accordion created:", class(acc), "\n")
if (exists("alpine_modal")) {
  cat("Modal component available\n")
}
if (exists("alpine_tabs")) {
  cat("Tabs component available\n")
}
```

# Pre-built Components and Patterns

This vignette covers alpinejs's pre-built interactive components: accordions, modals, and tabs. These are ready-to-use patterns built on top of Alpine.js.

## Accordion Component

An accordion displays collapsible panels where one or more can be open at a time.

### Basic Accordion

```{r}
alpine_accordion(
  items = list(
    list(title = "What is Alpine.js?", content = "Alpine.js is a lightweight, rugged JavaScript framework."),
    list(title = "Why use alpinejs?", content = "Write interactive UIs in pure R without JavaScript."),
    list(title = "Is it production-ready?", content = "Yes! Alpine.js is stable and widely used.")
  ),
  class = "faq-accordion"
)
```

### Accordion with Custom Styling

```{r}
# Styled accordion (uses inline CSS for demo)
alpine_accordion(
  items = list(
    list(
      title = "Section 1",
      content = result <- alpine_tag("div",
        style = "padding: 15px; background-color: #f5f5f5;",
        alpine_tag("p", "Content for section 1"),
        alpine_tag("ul",
          alpine_tag("li", "Point A"),
          alpine_tag("li", "Point B"),
          alpine_tag("li", "Point C")
        )
      )
    ),
    list(
      title = "Section 2",
      content = alpine_tag("div",
        style = "padding: 15px; background-color: #f5f5f5;",
        alpine_tag("p", "Content for section 2"),
        alpine_tag("code", "Some code goes here")
      )
    )
  ),
  class = "accordion",
  title_class = "accordion-title",
  content_class = "accordion-body"
)

cat("```html\n", as.character(result), "\n```", sep="")
```

### How It Works

The accordion uses Alpine.js internally:

```javascript
// Generated x-data state
{
  collapsed: [false, true, true],  // which panels are open
  togglePanel(index) { 
    this.collapsed[index] = !this.collapsed[index];
  }
}
```

Each panel header has `@click="togglePanel(index)"` and each body has `x-show="!collapsed[index]"`.

## Modal Component

A modal dialog overlays the page and captures user attention.

### Basic Modal

```{r}
alpine_modal(
  trigger_label = "Open Dialog",
  title = "Confirm Action",
  content = "Are you sure you want to proceed?"
)
```

### Modal with Rich Content

```{r}
# Modal with complex content
alpine_modal(
  trigger_label = "Learn More",
  title = "About Our Product",
  content = result <- alpine_tag("div",
    style = "line-height: 1.6;",
    alpine_tag("p", "Combines R and web technologies."),
    alpine_tag("h3", "Features:"),
    alpine_tag("ul",
      alpine_tag("li", "R interface to Alpine.js"),
      alpine_tag("li", "Type conversion"),
      alpine_tag("li", "Reactive bindings")
    ),
    alpine_tag("p",
      "Built with ",
      alpine_tag("a", "Alpine.js", href = "https://alpinejs.dev/", target = "_blank"),
      " and ",
      alpine_tag("a", "htmltools", href = "https://rstudio.github.io/htmltools/", target = "_blank")
    )
  ),
  close_label = "Got it!"
)

cat("```html\n", as.character(result), "\n```", sep="")
```

### Modal Without Trigger Button

For programmatic control (e.g., triggered after a server action):

```{r}
# Modal without automatic trigger
modal_component <- alpine_modal(
  trigger_label = NULL,  # No button
  title = "Processing...",
  content = "Your request is being processed. Please wait.",
  backdrop = FALSE  # No backdrop for a lighter feel
)

# Would be shown via JavaScript:
# modalOpen = true;
```

### Customizing Modal Appearance

The modal renders with basic inline styles. To customize, add CSS:

```css
/* Minimal custom modal styling */
.modal-dialog {
  box-shadow: 0 4px 6px rgba(0, 0, 0, 0.1);
  border-radius: 12px;
}

.modal-header {
  border-bottom: 1px solid #e5e7eb;
  padding-bottom: 10px;
}

.modal-footer {
  border-top: 1px solid #e5e7eb;
  padding-top: 15px;
}
```

## Transitions (x-transition)

`alpine_transition()` adds enter/leave animations to any element controlled by
`alpine_show()`. Two modes are available: a built-in transition helper (fade + scale)
and CSS class mode for full control with utility frameworks like Tailwind.

### Built-in Transition Helper

```{r}
# Default fade + scale transition
alpine_tag("div",
  alpine_tag("button", "Toggle",
    alpine_attr(`@click` = "open = !open")
  ),
  alpine_tag("div",
    style = "padding: 10px; background: #e8f4fd; margin-top: 8px;",
    "This content fades in and out."
  ) |>
    alpine_show("open") |>
    alpine_transition()
) |>
  alpine_data(open = FALSE)
```

### Custom Duration and Opacity Only

```{r}
# Opacity-only fade, 400ms
alpine_tag("div",
  alpine_tag("button", "Fade",
    alpine_attr(`@click` = "visible = !visible")
  ),
  alpine_tag("p", "Fades smoothly with no scale effect.") |>
    alpine_show("visible") |>
    alpine_transition(.opacity = TRUE, .duration = 400)
) |>
  alpine_data(visible = TRUE)
```

### Scale Transition with Custom Origin

```{r}
# Scale from top, 300ms
alpine_tag("div",
  alpine_tag("button", "Open Dropdown",
    alpine_attr(`@click` = "open = !open")
  ),
  alpine_tag("ul",
    style = "border: 1px solid #ccc; padding: 8px;",
    alpine_tag("li", "Option A"),
    alpine_tag("li", "Option B"),
    alpine_tag("li", "Option C")
  ) |>
    alpine_show("open") |>
    alpine_transition(.scale = TRUE, .scale_value = 90, .origin = "top", .duration = 200)
) |>
  alpine_data(open = FALSE)
```

### CSS Class Mode (Tailwind-style)

For full control over transition phases, provide CSS class strings for each phase.
This matches how Tailwind CSS transitions are structured:

```{r, eval=FALSE}
# CSS class mode -- attach separate x-transition:phase="classes" attributes
alpine_tag("div", "Slides in") |>
  alpine_show("open") |>
  alpine_transition(
    .enter       = "transition ease-out duration-300",
    .enter_start = "opacity-0 translate-x-4",
    .enter_end   = "opacity-100 translate-x-0",
    .leave       = "transition ease-in duration-200",
    .leave_start = "opacity-100 translate-x-0",
    .leave_end   = "opacity-0 translate-x-4"
  )
```

Note: CSS class mode requires the CSS classes to be defined in your stylesheet
(e.g., via Tailwind CSS or custom classes). The helper mode works with no extra CSS.

## Tabs Component

### Basic Tabs

```{r}
alpine_tabs(
  tabs = list(
    list(label = "Overview", content = "This is the overview tab"),
    list(label = "Details", content = "Detailed information goes here"),
    list(label = "Contact", content = "Contact us for more information")
  )
)
```

### Tabs with Rich Content

```{r}
# Tabs with complex content
alpine_tabs(
  tabs = list(
    list(
      label = "Home",
      content = result <- alpine_tag("div",
        alpine_tag("h3", "Welcome"),
        alpine_tag("p", "This is the home tab content."),
        alpine_tag("button", "Learn More",
          alpine_attr(`@click` = "alert('More info')")
        )
      )
    ),
    list(
      label = "Documentation",
      content = alpine_tag("div",
        alpine_tag("h3", "Docs"),
        alpine_tag("p", "Read the documentation here."),
        alpine_tag("ul",
          alpine_tag("li", "Installation"),
          alpine_tag("li", "Getting started"),
          alpine_tag("li", "API reference")
        )
      )
    ),
    list(
      label = "Examples",
      content = alpine_tag("div",
        alpine_tag("h3", "Code Examples"),
        alpine_tag("p", "View working examples below."),
        alpine_tag("pre",
          alpine_tag("code", "# Example R code\nalpine_tag(\"div\") |> alpine_data(x = 1)")
        )
      )
    )
  ),
  class = "custom-tabs",
  tab_class = "custom-tab-button",
  content_class = "custom-tab-panel"
)

cat("```html\n", as.character(result), "\n```", sep="")
```

### How Tabs Work

The tabs component uses:

```javascript
// Generated x-data state
{
  activeTab: 0  // 0-indexed active tab
}
```

Each tab button has `@click="activeTab = index"` and each content panel has `x-show="activeTab === index"`.

Tab buttons get dynamic class binding via `x-bind:class` to highlight the active tab.

## Combining Components

Nest and combine components together:

```{r}
# Tabs containing accordions
page <- result <- alpine_tag("div",
  alpine_tag("h1", "Documentation"),
  
  alpine_tabs(
    tabs = list(
      list(
        label = "Getting Started",
        content = alpine_accordion(
          items = list(
            list(title = "Installation", content = "Install alpinejs from GitHub..."),
            list(title = "First Steps", content = "Start by loading the library...")
          )
        )
      ),
      list(
        label = "API Reference",
        content = alpine_accordion(
          items = list(
            list(title = "alpine_data()", content = "Initialize component state..."),
            list(title = "alpine_on()", content = "Attach event listeners...")
          )
        )
      )
    )
  )
)

page

cat("```html\n", as.character(result), "\n```", sep="")
```

## Component Composition Pattern

Create reusable component functions:

```{r}
# Reusable confirmation dialog
confirm_dialog <- function(message, trigger_label = "Confirm") {
  alpine_modal(
    trigger_label = trigger_label,
    title = "Confirmation",
    content = message,
    close_label = "Cancel"
  )
}

# Use it multiple times with distinct button labels
result <- alpine_tag("div",
  confirm_dialog("Delete this item?", "Delete"),
  confirm_dialog("Log out?", "Log Out")
)

cat("```html\n", as.character(result), "\n```", sep="")
```

## Advanced: Data-Driven Components

Create components from data:

```{r}
# Build an accordion from a data frame
faq_data <- data.frame(
  question = c("What is R?", "What is Alpine.js?", "What is alpinejs?"),
  answer = c(
    "R is a language for statistical computing...",
    "Alpine.js is a lightweight JavaScript framework...",
    "alpinejs brings Alpine.js to R..."
  ),
  stringsAsFactors = FALSE
)

# Convert to list format expected by alpine_accordion
items <- lapply(seq_len(nrow(faq_data)), function(i) {
  list(
    title = faq_data$question[i],
    content = faq_data$answer[i]
  )
})

alpine_accordion(items)
```

## Styling Best Practices

### With Tailwind CSS

```r
# Accordion with Tailwind classes
alpine_accordion(
  items = list(
    list(title = "Item 1", content = "Content 1"),
    list(title = "Item 2", content = "Content 2")
  ),
  class = "divide-y divide-gray-200 border border-gray-200 rounded-lg",
  title_class = "px-4 py-3 font-semibold hover:bg-gray-50 cursor-pointer",
  content_class = "px-4 py-2 bg-gray-50"
)

cat("```html\n", as.character(.), "\n```", sep="")

```

### With Bootstrap

```r
# Modal with Bootstrap styling
alpine_modal(
  trigger_label = "Open",
  title = "Bootstrap Modal",
  content = "Content styled with Bootstrap classes",
  class = "modal-lg"
)

cat("```html\n", as.character(.), "\n```", sep="")

```

## Performance Notes

- **Lightweight**: All components are pure Alpine.js--no heavy dependencies
- **No Server Calls**: Component state is managed client-side
- **Fast Rendering**: HTML is generated once; only state changes are reactive
- **Suitable for**: 100s to 1000s of items; for larger datasets, consider server-side pagination

## Accessibility

alpinejs components use semantic HTML. For better accessibility:

```r
# Add ARIA labels
alpine_accordion(
  items = list(
    list(title = "Section 1", content = "Content 1")
  ),
  alpine_attr(`x-bind:aria-expanded` = "true")  # Make dynamic for each panel
)

cat("```html\n", as.character(.), "\n```", sep="")

```

## When to Use Pre-built Components

**Good for:**

- Modals, dialogs, confirmations
- Tabbed interfaces
- Collapsible sections/FAQs
- Mobile menus/dropdowns

**Consider alternatives for:**

- Large tables (use `DT` + Shiny)
- Real-time data (use WebSocket + other frameworks)
- Complex workflows (use Shiny app)

## Next Steps

- **Reactivity and Global State**: See the [Reactivity and Globals Vignette](reactivity-and-globals.qmd) for `$watch`, `$dispatch`, and `Alpine.store()`
- **Advanced Patterns**: Learn about custom components and composition
- **Crosstalk Integration**: Filter widgets with Alpine controls

- **Performance Optimization**: Techniques for large-scale applications

---

For more examples, see the `inst/examples/` directory in the alpinejs package or run examples with `Rscript inst/examples/01-todo-list.R`.
