---
title: Getting Started with alpinejs
---


```{r, include = FALSE
litedown::reactor(
  collapse = TRUE,
  comment = "#> "
)
```

## Installation

```{r, eval=FALSE}
# Install from GitHub
remotes::install_github("brownag/alpinejs")
```

```{r setup, eval=TRUE, echo=TRUE}
library(alpinejs)
library(htmltools)
```

# Getting Started with alpinejs

The **alpinejs** R package lets you build interactive HTML applications without writing JavaScript. You write R code; alpinejs generates the HTML and JavaScript that Alpine.js (a lightweight JavaScript framework) uses to handle interactivity in the browser.

```{r test_basic_concepts, eval=TRUE, echo=FALSE}
# Verify core functions work and return proper tags
result <- div() |> alpine_data(count = 0)
cat("alpine_data creates class:", class(result), "\n")
cat("Has Alpine attribute:", "x-data" %in% names(attributes(result)), "\n")
```

## What is alpinejs?

alpinejs is an R package that:

1. **Generates HTML** using R functions like `alpine_tag()`, `alpine_data()`, `alpine_on()`, etc.
2. **Creates Alpine.js directives** inside that HTML (attributes like `x-data`, `@click`, `x-model`)
3. **Serializes R data to JSON** for Alpine.js to use as reactive state
4. **Works with R's pipe operator** (`|>`) for clean, readable chains

In the browser, Alpine.js then handles reactivity, event listeners, and DOM updates. You never write JavaScript—only R!

## Core Concepts

### State (x-data)

Every Alpine component starts with **data** -- the reactive state that drives the UI:

```{r}
alpine_tag("div",
  alpine_tag("button", "Count: ", alpine_tag("span",
    alpine_attr(`x-bind:textContent` = "count")
  ))
) |>
  alpine_data(count = 0)
```

The `alpine_data()` function initiates a component's state. Pass named arguments to set initial values:

```{r}
alpine_tag("div",
  alpine_tag("p", "Username: ", alpine_tag("span",
    alpine_attr(`x-bind:textContent` = "username")
  )),
  alpine_tag("p", "Email: ", alpine_tag("span",
    alpine_attr(`x-bind:textContent` = "email")
  )),
  alpine_tag("p", "Logged in: ", alpine_tag("span",
    alpine_attr(`x-bind:textContent` = "isLoggedIn")
  ))
) |>
  alpine_data(
    username = "Alice",
    email = "alice@example.com",
    isLoggedIn = TRUE
  )
```

### Events (x-on)

Respond to user interactions with `alpine_on()`:

```{r}
alpine_tag("div",
  alpine_tag("button", "Increment",
    alpine_attr(`@click` = "count++")
  ),
  alpine_tag("p", "Count: ", alpine_tag("span",
    alpine_attr(`x-bind:textContent` = "count")
  ))
) |>
  alpine_data(count = 0)
```

Common events:

- **Mouse**: `click`, `dblclick`, `mouseenter`, `mouseleave`
- **Keyboard**: `keydown`, `keyup`, `keypress`
- **Form**: `input`, `change`, `submit`, `focus`, `blur`

### Conditional Display (x-show)

Show and hide elements based on boolean conditions:

```{r}
alpine_tag("div",
  alpine_tag("button", "Toggle Details",
    alpine_attr(`@click` = "showDetails = !showDetails")
  ),
  alpine_tag("div",
    "These are secret details!",
    style = "color: red; margin-top: 10px;",
    alpine_attr(`x-show` = "showDetails")
  )
) |>
  alpine_data(showDetails = FALSE)
```

### Conditional Rendering (x-if)

`alpine_if()` completely removes an element from the DOM when the condition is false.
Use it for elements that are rarely shown or that are expensive to initialize.

```{r}
alpine_tag("div",
  alpine_tag("button", "Show Admin Panel",
    alpine_attr(`@click` = "isAdmin = !isAdmin")
  ),
  alpine_if(
    alpine_tag("div",
      style = "padding: 10px; background: #fee;",
      "Admin controls here"
    ),
    "isAdmin"
  )
) |>
  alpine_data(isAdmin = FALSE)
```

**`x-show` vs `x-if`:** Use `alpine_show()` (CSS hide) for elements that toggle
frequently or need state preserved between toggles. Use `alpine_if()` (DOM removal)
for elements that are rarely shown, conditionally loaded content, or when you want
to guarantee the element's lifecycle hooks run on each toggle.

### Preventing Flash of Unrendered Content (x-cloak)

When Alpine.js loads asynchronously, reactive elements may briefly flash their raw
template state. `alpine_cloak()` hides elements until Alpine has initialized them.

```{r}
alpine_tag("div",
  alpine_tag("span") |>
    alpine_text("message") |>
    alpine_cloak()
) |>
  alpine_data(message = "Hello from Alpine!")
```

`alpine_cloak()` automatically injects the required CSS (`[x-cloak] { display: none !important; }`)
so no additional stylesheet configuration is needed.

### DOM References (x-ref and $refs)

Label elements with `alpine_ref()` and access them directly in event handlers
using `$refs.name`. This replaces manual `getElementById` or `querySelector` calls.

```{r}
alpine_tag("div",
  alpine_tag("input", type = "text", placeholder = "Type something...") |>
    alpine_ref("textInput"),
  alpine_tag("button", "Copy to Clipboard") |>
    alpine_on("click", "navigator.clipboard.writeText($refs.textInput.value)")
) |>
  alpine_data(x = "")
```

Use `alpine_refs()` to generate `$refs.name` expressions for use in other directives:

```{r, eval=FALSE}
# Programmatic $refs expression (useful when building complex expressions)
alpine_refs("textInput")
# Returns JS expression: $refs.textInput
```

### Two-Way Binding (x-model)

Synchronize input fields with state:

```{r}
alpine_tag("div",
  alpine_tag("input", type = "text", placeholder = "Enter your name") |> alpine_model("name"),
  alpine_tag("p", "Hello, ", alpine_tag("span",
    alpine_attr(`x-bind:textContent` = "name")
  ), "!")
) |>
  alpine_data(name = "")
```

Typing in the input field automatically updates the `name` state, and the greeting updates reactively.

### Attribute Binding (x-bind)

Dynamically set HTML attributes:

```{r}
alpine_tag("div",
  alpine_tag("input",
    type = "text",
    placeholder = "Enter username",
    alpine_attr(`x-model` = "username")
  ),
  alpine_tag("button", "Submit",
    alpine_attr(`x-bind:disabled` = "username.length === 0")
  )
) |>
  alpine_data(username = "")
```

The button is disabled until the username field has text. Note how we use `alpine_attr()` to keep HTML attributes and Alpine directives organized together.

## Practical Example: A Form with Validation

Build a form that validates in real-time:

```{r}
alpine_tag("div",
  style = "max-width: 400px; margin: 20px auto;",
  alpine_tag("h2", "Register"),
  
  alpine_tag("label", "Full Name:", style = "display: block; margin-top: 10px; font-weight: bold;"),
  alpine_tag("input", type = "text", style = "width: 100%; padding: 5px;") |> alpine_model("fullName"),
  
  alpine_tag("label", "Email:", style = "display: block; margin-top: 10px; font-weight: bold;"),
  alpine_tag("input", type = "email", style = "width: 100%; padding: 5px;") |> alpine_model("email"),
  
  alpine_tag("label", "Password:", style = "display: block; margin-top: 10px; font-weight: bold;"),
  alpine_tag("input", type = "password", style = "width: 100%; padding: 5px;") |> alpine_model("password"),
  
  # Validation messages
  alpine_tag("div",
    "Email is invalid",
    style = "color: red; margin-top: 5px; font-size: 12px;"
  ) |> alpine_show("email.length > 0 && !isValidEmail()"),
  
  alpine_tag("div",
    "Password must be at least 8 characters",
    style = "color: red; margin-top: 5px; font-size: 12px;"
  ) |> alpine_show("password.length > 0 && password.length < 8"),
  
  # Submit button
  alpine_tag("button",
    "Register",
    style = "margin-top: 20px; padding: 10px 20px;",
    alpine_attr(`x-bind:disabled` = "!isFormValid()"),
    alpine_attr(`@click` = "submitForm()")
  )
) |>
  alpine_data(
    fullName = "",
    email = "",
    password = "",
    isValidEmail = htmlwidgets::JS(
      "function() { return /^[^\\s@]+@[^\\s@]+\\.[^\\s@]+$/.test(this.email); }"
    ),
    isFormValid = htmlwidgets::JS(
      "function() { return this.fullName.length > 0 && this.isValidEmail() && this.password.length >= 8; }"
    ),
    submitForm = htmlwidgets::JS(
      "function() { alert('Form submitted!'); }"
    )
  )
```

This example demonstrates:

- Multiple state variables
- Real-time validation
- Conditional showing of error messages
- JavaScript functions injected with `htmlwidgets::JS()`
- Dynamic button disabling

## Data Type Conversions

alpinejs automatically converts R data types to JSON. The conversion is as follows:

```{r}
# Various data types
alpine_tag("div", "Data demo") |>
  alpine_data(
    # Simple types
    count = 42,
    name = "Alice",
    isActive = TRUE,
    score = 3.14,
    
    # Collections
    colors = c("red", "blue", "green"),
    metadata = list(version = "1.0", author = "dev"),
    
    # Factors (converted to character)
    category = factor(c("A", "B", "C")),
    
    # Dates (converted to ISO 8601)
    today = Sys.Date(),
    now = Sys.time(),
    
    # NA becomes null
    missing = NA
  ) |>
  htmltools::browsable()
```

Result in Alpine state:
```javascript
{
  count: 42,
  name: "Alice",
  isActive: true,
  score: 3.14,
  colors: ["red", "blue", "green"],
  metadata: {version: "1.0", author: "dev"},
  category: ["A", "B", "C"],
  today: "2026-02-21",
  now: "2026-02-21T10:30:45Z",
  missing: null
}
```

## JavaScript Expressions

For custom interactivity, inject raw JavaScript using `htmlwidgets::JS()`:

```{r}
alpine_tag("div",
  alpine_tag("button", "Click me",
    alpine_attr(`@click` = "greet()")
  ),
  alpine_tag("p",
    alpine_attr(`x-bind:textContent` = "greeting")
  )
) |>
  alpine_data(
    greeting = "Hello!",
    greet = htmlwidgets::JS("function() { this.greeting = 'You clicked!'; }")
  )
```

**Important:** CSP mode (default) has limitations on JavaScript expressions. Stick to:
- Simple variable assignments
- Method calls
- Basic operators
- Conditionals

Avoid closures, function definitions, or complex expressions in CSP mode.

## Exporting to HTML Files

Generate standalone HTML files that work without a server:

```{r}
alpine_tag("div",
  alpine_tag("button", "Toggle") |> alpine_on("click", "show = !show"),
  alpine_tag("div", "Hidden content") |> alpine_show("show")
) |>
  alpine_data(show = FALSE)
```

## Integration with Shiny

Use alpinejs components in Shiny for lightweight interactivity:

```{r, eval=FALSE}
library(shiny)
library(alpinejs)

ui <- fluidPage(
  alpine_tag("h1", "My App"),
  
  alpine_tag("div",
    alpine_tag("input", type = "text",
      alpine_attr(`x-model` = "text")
    ),
    alpine_tag("p", "You typed: ", alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "text")
    ))
  ) |>
    alpine_data(text = "")
)

server <- function(input, output, session) {
  # Server logic here
}

shinyApp(ui, server)
```

## Next Steps

- **Pre-built Components**: See the [Components Vignette](components-and-patterns.qmd) for accordion, modal, and tabs examples
- **Reactivity and Global State**: See the [Reactivity and Globals Vignette](reactivity-and-globals.qmd) for `$watch`, `$dispatch`, and `Alpine.store()`
- **Advanced Patterns**: Learn about composition, nested components, and performance optimization
- **Deeper Integration**: Custom event handlers and component composition patterns

## Best Practices

1. **Keep State Simple**: Use Alpine for UI logic, not business logic
2. **Use Events Sparingly**: For simple interactions; use HTMX/Plumber for server calls
3. **Validate Client-Side**: Show errors early with `x-show`
4. **Structure Components**: Create reusable R functions that return Alpine components
5. **Test Rendered HTML**: Verify the final HTML structure in browser dev tools

## Troubleshooting

**Q: My Alpine component isn't interactive**

- Check browser console for JavaScript errors
- Verify CDN is loaded (should see `<script src="...alpinejs..."></script>` in page source)
- Ensure Alpine expressions are valid JavaScript

**Q: State isn't updating**

- Verify state variable names match in `x-data` and expressions
- Check that event handlers are spelled correctly (`x-on`, not `@on`)
- Use browser DevTools to inspect the Alpine component state

**Q: CSP security warning?**

- alpinejs uses CSP-compliant build by default
- Some complex JavaScript may require `options(alpinejs.use_csp = FALSE)`
- Document any CSP exceptions in your app's policy

---

See the next vignette for pre-built components!
