---
title: Input Modifiers: Enhancing Form Controls
---



```{r setup, include=FALSE}
litedown::reactor(echo = TRUE)
RUN_EXAMPLES <- nzchar(Sys.getenv("R_ALPINEJS_RUN_EXAMPLES"))
library(alpinejs)
library(htmltools)
```

```{r test_modifiers, eval=RUN_EXAMPLES, echo=FALSE}
# Verify modifiers work and return proper structure
result <- alpine_tag("input") |> alpine_model("field", .lazy = TRUE)
cat("alpine_model with .lazy creates proper tag\n")
cat("Class:", class(result), "\n")
```

## Overview

Alpine.js input modifiers modify form control behavior including debouncing, lazy updates, and automatic type conversion. The `alpine_model()` function supports these modifiers using parameters prefixed with a dot (`.`).

**Available Modifiers:**

| Modifier | Behavior | Use Case |
|----------|----------|----------|
| `.lazy` | Update on blur instead of keystroke | Validation, expensive operations |
| `.debounce` | Delay updates by N milliseconds | Search, filter, API calls |
| `.number` | Auto-convert input to numeric | Age, price, quantity fields |
| `.trim` | Strip whitespace | Username, email, API keys |

## Basic Usage

### Syntax

```r
# Single modifier
alpine_tag("input") |> alpine_model("fieldName", .lazy = TRUE)

# Multiple modifiers
alpine_tag("input") |> alpine_model("fieldName", .lazy = TRUE, .debounce = 300, .number = TRUE)

# Pipe-friendly chaining
alpine_tag("input", placeholder = "Search...") |>
  alpine_model("query", .debounce = 300)
```

```{r verify_modifiers, eval=RUN_EXAMPLES, echo=FALSE}
# Test all modifiers work and create proper tags
lazy_input <- alpine_tag("input") |> alpine_model("field1", .lazy = TRUE)
debounce_input <- alpine_tag("input") |> alpine_model("field2", .debounce = 300)
number_input <- alpine_tag("input") |> alpine_model("field3", .number = TRUE)
trim_input <- alpine_tag("input") |> alpine_model("field4", .trim = TRUE)
combined <- alpine_tag("input") |> alpine_model("field5", .lazy = TRUE, .debounce = 500, .trim = TRUE)

cat("All modifiers work successfully:\n")
cat("- .lazy\n")
cat("- .debounce\n")
cat("- .number\n")
cat("- .trim\n")
cat("- Combined modifiers\n")
```

### Simple Example

```r
library(alpinejs)
library(htmltools)

alpine_tag("div",
  alpine_tag("label", "Email:"),
  alpine_tag("input",
    type = "email",
    placeholder = "Enter email"
  ) |>
    alpine_model("email", .lazy = TRUE),
  
  alpine_tag("div",
    "Current: ", 
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "email")
    )
  )
) |>
  alpine_data(email = "") |>
  htmltools::browsable()
```

The `.lazy` modifier defers updates until the element loses focus (blur event).

## Modifier Behaviors

### 1. `.lazy = TRUE` -- Lazy Evaluation

By default, Alpine.js updates the model on every keystroke. The `.lazy` modifier defers updates until the element loses focus (blur event).

**When to use:**

- Form validation (email, password, username)
- Expensive operations (database lookups)
- Reducing network traffic

```r
# Email validation example
alpine_tag("div",
  alpine_tag("input",
    type = "email",
    placeholder = "your@email.com"
  ) |>
    alpine_model("email", .lazy = TRUE),
  
  alpine_tag("span",
    "Valid: ",
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "isValidEmail()")
    )
  )
) |>
  alpine_data(
    email = "",
    isValidEmail = htmlwidgets::JS(
      "function() { return this.email.includes('@'); }"
    )
  )
```

### 2. `.debounce = N` -- Debounced Updates

Debouncing adds a delay (in milliseconds) between keystroke and model update.

**When to use:**

- Search inputs (typing triggers database queries)
- Auto-save features
- Real-time filtering
- API rate limiting

```r
# Live search example
alpine_tag("div",
  alpine_tag("input",
    type = "text",
    placeholder = "Search..."
  ) |>
    alpine_model("searchQuery", .debounce = 300),
  
  alpine_tag("ul",
    alpine_tag("li", "Results will appear here")
  )
) |>
  alpine_data(searchQuery = "") |>
  htmltools::browsable()

# With debounce = 300ms, updates trigger 300ms after user stops typing
```

Typical use: 200-500ms.

### 3. `.number = TRUE` -- Automatic Type Conversion

Type conversion converts string input to numeric value.

**When to use:**

- Numeric form fields (age, price, quantity)
- Math operations
- Range sliders, spinners

```r
# Age input example
alpine_tag("div",
  alpine_tag("label", "Age:"),
  alpine_tag("input",
    type = "number",
    min = "0",
    max = "120",
    placeholder = "Enter age"
  ) |>
    alpine_model("age", .number = TRUE),
  
  alpine_tag("div",
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "age")
    ),
    " years old"
  )
) |>
  alpine_data(age = 0) |>
  htmltools::browsable()
```

**Automatic behavior:**

- Input `"25"` -> `age = 25` (numeric)
- Input `"invalid"` -> `age = NaN`
- Input `""` -> `age = null`

### 4. `.trim = TRUE` -- Strip Whitespace

Removes leading and trailing whitespace before updating the model.

**When to use:**

- Username/email fields
- API keys, credentials
- Search inputs

```r
# Username input
alpine_tag("div",
  alpine_tag("input",
    placeholder = "Username"
  ) |>
    alpine_model("username", .trim = TRUE),
  
  alpine_tag("div",
    "Length: ",
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "username.length")
    )
  )
) |>
  alpine_data(username = "") |>
  htmltools::browsable()
```

**Behavior:**

- Input `"  john  "` -> `username = "john"`
- Input `"  a  "` -> `username = "a"`

## Combining Modifiers

Multiple modifiers work together:

```r
# Form field with all modifiers
alpine_tag("textarea", placeholder = "Notes...") |>
  alpine_model(
    "notes",
    .lazy = TRUE,      # Update on blur
    .debounce = 500,   # Wait 500ms
    .trim = TRUE       # Remove whitespace
  )

# This updates only after user:
# 1. Stops typing for 500ms, AND
# 2. Clicks away, AND
# 3. The value gets trimmed
```

## Real-World Examples

### Example 1: Password Strength Validator

```r
library(alpinejs)
library(htmltools)

alpine_tag("div",
  style = "max-width: 400px; padding: 20px;",
  
  alpine_tag("h3", "Password Strength Checker"),
  
  # Password input
  alpine_tag("div",
    alpine_tag("label", "Password:"),
    alpine_tag("input",
      type = "password",
      placeholder = "Enter password"
    ) |>
      alpine_model("password", .lazy = TRUE)
  ),
  
  # Strength indicator
  alpine_tag("div",
    style = "display: flex; gap: 5px; margin-top: 10px;",
    alpine_tag("div",
      style = "width: 25px; height: 4px; background: #ccc;",
      alpine_attr(`x-bind:class` = "strengthClass()")
    ),
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "strengthLabel()")
    )
  )
) |>
  alpine_data(
    password = "",
    strengthClass = htmlwidgets::JS("function() {
      const len = this.password.length;
      if (len < 4) return 'weak';
      if (len < 8) return 'fair';
      if (len < 12) return 'good';
      return 'excellent';
    }"),
    strengthLabel = htmlwidgets::JS("function() {
      const classes = {weak: 'Weak', fair: 'Fair', good: 'Good', excellent: 'Excellent'};
      return classes[this.strengthClass()];
    }")
  ) |>
  htmltools::browsable()
```

### Example 2: Dynamic Form with Debounced Search

```r
library(alpinejs)
library(htmltools)

alpine_tag("div",
  style = "padding: 20px;",
  
  alpine_tag("h3", "User Search"),
  
  # Search input with .debounce
  alpine_tag("input",
    type = "text",
    placeholder = "Search users..."
  ) |>
    alpine_model("searchQuery", .debounce = 300),
  
  # Loading indicator
  alpine_tag("div",
    "Loading..." |>
      alpine_show("isSearching")
  ),
  
  # Results
  alpine_tag("div",
    alpine_tag("ul",
      alpine_tag("li",
        "Result 1" |> alpine_show("!isSearching && searchQuery")
      )
    )
  )
) |>
  alpine_data(
    searchQuery = "",
    isSearching = htmlwidgets::JS("false"),
    performSearch = htmlwidgets::JS("
      function() {
        if (!this.searchQuery) return;
        this.isSearching = true;
        // Simulate API call
        setTimeout(() => { this.isSearching = false; }, 500);
      }
    ")
  ) |>
  htmltools::browsable()
```

### Example 3: Multi-Field Form with Mixed Modifiers

```r
# Registration form with different modifier strategies
alpine_tag("form",
  style = "max-width: 500px; padding: 20px;",
  
  # Username: lazy + trim
  alpine_tag("div",
    alpine_tag("label", "Username:"),
    alpine_tag("input",
      type = "text",
      placeholder = "Choose username"
    ) |>
      alpine_model("username", .lazy = TRUE, .trim = TRUE)
  ),
  
  # Email: lazy (for validation)
  alpine_tag("div",
    alpine_tag("label", "Email:"),
    alpine_tag("input",
      type = "email",
      placeholder = "your@email.com"
    ) |>
      alpine_model("email", .lazy = TRUE)
  ),
  
  # Age: number conversion
  alpine_tag("div",
    alpine_tag("label", "Age:"),
    alpine_tag("input",
      type = "number",
      min = "18",
      max = "120"
    ) |>
      alpine_model("age", .number = TRUE)
  ),
  
  # Bio: lazy + debounce + trim (auto-save)
  alpine_tag("div",
    alpine_tag("label", "Bio:"),
    alpine_tag("textarea", placeholder = "Tell us about yourself") |>
      alpine_model("bio", .lazy = TRUE, .debounce = 500, .trim = TRUE)
  ),
  
  # Submit button
  alpine_tag("button", "Register") |>
    alpine_on("click", "submitForm()")
) |>
  alpine_data(
    username = "",
    email = "",
    age = 18,
    bio = "",
    submitForm = htmlwidgets::JS("function() {
      console.log({
        username: this.username,
        email: this.email,
        age: this.age,
        bio: this.bio
      });
    }")
  ) |>
  htmltools::browsable()
```

## Browser Compatibility

All input modifiers work in modern browsers:

- Chrome/Edge 88+
- Firefox 78+
- Safari 14+
- IE: Not supported (use polyfill if needed)

## Performance Optimization

### Debounce Values by Use Case

| Use Case | Recommended | Rationale |
|----------|-------------|-----------|
| Live search | 300-500ms | Balance responsiveness & server load |
| Input validation | 200ms | Fast feedback for users |
| Text input | 150-300ms | Feel responsive without lag |
| Multi-field form | 200-400ms | Depends on complexity |
| Auto-save | 500ms-1s | Reduces API calls |

### Common Pattern: Search with Loading State

```r
# Optimize with debounce + loading indicator
alpine_tag("div",
  alpine_tag("input",
    type = "text",
    placeholder = "Search..."
  ) |>
    alpine_model("query", .debounce = 400),  # Wait 400ms
  
  alpine_tag("div", "Searching...") |> alpine_show("isSearching")
) |>
  alpine_data(
    query = "",
    isSearching = htmlwidgets::JS("false")
  )
```

## Troubleshooting

**Q: Input seems unresponsive with .lazy modifier**
A: This is normal--updates only happen on blur. Combine with `.debounce` for faster feedback during typing.

**Q: Numbers appearing as strings despite .number**
A: Verify input uses `type = "number"` and check browser console for errors.

**Q: Debounce delays too much**
A: Reduce the delay value (try 200ms instead of 500ms) or combine with other feedback mechanisms.

**Q: Whitespace not trimming with .trim**
A: Trim only applies to model binding, not to display. Trim in JavaScript if needed: `value.trim()`.

## See Also

- [Alpine.js Modifier Documentation](https://alpinejs.dev/)
- [Getting Started with alpinejs](getting-started.qmd)
- [Components and Patterns](components-and-patterns.qmd)
