---
title: Performance & Best Practices
---



```{r setup, include=FALSE}
litedown::reactor(echo = TRUE)
RUN_EXAMPLES <- nzchar(Sys.getenv("R_ALPINEJS_RUN_EXAMPLES"))
library(alpinejs)
```

```{r test_performance, eval=RUN_EXAMPLES, echo=FALSE}
# Verify performance patterns work
cat("Creating different DOM structures to test\n")
# Both slow and fast versions should build valid tags
slow <- alpine_tag("ul",
  alpine_tag("li", "item 1"), 
  alpine_tag("li", "item 2")
)
fast <- alpine_tag("ul",
  alpine_tag("li", "item 1"), 
  alpine_tag("li", "item 2")
)
cat("Both structures created successfully\n")
```

## Overview

This vignette covers optimization strategies, debugging techniques, and best practices for building performant, maintainable Alpine.js applications in R.

## Performance Optimization

### 1. Reduce DOM Size

Minimize the number of DOM nodes, especially in loops:

```r
# Slow approach: Renders every field for every item
alpine_tag("ul",
  alpine_tag("li",
    ":for" = "item in items",
    # 20 child elements per item
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "item.name")
    ),
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "item.email")
    ),
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "item.status")
    )
    # ... many more spans
  )
)

# Faster approach: Use single computed property
alpine_tag("ul",
  alpine_tag("li",
    ":for" = "item in items",
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "formatItem(item)")
    )
  )
) |>
  alpine_data(
    items = list(...),
    formatItem = htmlwidgets::JS("
      function(item) {
        return item.name + ' (' + item.email + ')';
      }
    ")
  )
```

### 2. Use Computed Properties Strategically

Pre-compute expensive operations instead of recalculating on every render:

```r
# Slow approach: Recalculates on every update
alpine_tag("div",
  alpine_tag("span",
    alpine_attr(`x-bind:textContent` = "items.filter(i => i.active).length")
  )
)

# Faster approach: Computed property with caching
alpine_tag("div",
  alpine_tag("span",
    alpine_attr(`x-bind:textContent` = "activeCount")
  )
) |>
  alpine_data(
    items = list(...),
    activeCount = htmlwidgets::JS("
      get activeCount() {
        return this.items.filter(i => i.active).length;
      }
    ")
  )
```

### 3. Debounce Expensive Operations

Use `.debounce` modifier for search, filter, and validation:

```r
# Slow approach: API call on every keystroke
alpine_tag("input") |>
  alpine_model("query") |>
  alpine_on("change", "performSearch()")

# Faster approach: Debounced search (wait 300ms)
alpine_tag("input") |>
  alpine_model("query", .debounce = 300) |>
  alpine_on("change", "performSearch()")
```

### 4. Limit Loop Rendering with Pagination

Large datasets should be paginated client-side or server-side:

```r
# Recommended approach: Show only current page
alpine_tag("div",
  alpine_tag("ul",
    alpine_tag("li",
      ":for" = "(item, idx) in visibleItems",
      alpine_tag("span",
        alpine_attr(`x-bind:textContent` = "item.name")
      )
    )
  ),
  
  # Pagination controls
  alpine_tag("button", "Previous") |>
    alpine_on("click", "page--"),
  alpine_tag("span",
    alpine_attr(`x-bind:textContent` = "page")
  ),
  alpine_tag("button", "Next") |>
    alpine_on("click", "page++")
) |>
  alpine_data(
    items = list(...),  # 10,000 items
    page = 1,
    pageSize = 20,
    visibleItems = htmlwidgets::JS("
      get visibleItems() {
        const start = (this.page - 1) * this.pageSize;
        return this.items.slice(start, start + this.pageSize);
      }
    ")
  )
```

### 5. Avoid Excessive Event Listeners

Attach listeners at parent level (event delegation):

```r
# Slow approach: Separate listener for each item
alpine_tag("ul",
  alpine_tag("li",
    ":for" = "item in items",
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "item.name")
    ),
    alpine_tag("button", "Delete") |>
      alpine_on("click", "deleteItem(item.id)")
  )
)

# Faster approach: Single parent listener with event delegation
alpine_tag("ul",
  alpine_tag("li",
    ":for" = "item in items",
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "item.name")
    ),
    alpine_tag("button", ":data-id" = "item.id", "Delete")
  ) |>
    alpine_on("click", "handleDelete(event)")  # Single listener
) |>
  alpine_data(
    items = list(...),
    handleDelete = htmlwidgets::JS("
      function(e) {
        if (e.target.tagName === 'BUTTON') {
          this.deleteItem(e.target.dataset.id);
        }
      }
    ")
  )
```

## Data & State Management

### 1. Structure State Logically

Group related data in nested objects:

```r
# Poor organization: Flat structure
alpine_data(
  username = "",
  email = "",
  phone = "",
  address = "",
  city = "",
  country = ""
  # ... many more fields
)

# Better organization: Nested objects
alpine_data(
  user = htmlwidgets::JS("
    {
      username: '',
      email: '',
      phone: ''
    }
  "),
  address = htmlwidgets::JS("
    {
      street: '',
      city: '',
      country: ''
    }
  ")
)

# Access as: this.user.email, this.address.city
```

### 2. Use Proper Data Types

Match JavaScript types to your data:

```r
# Incorrect: Numbers as strings
alpine_data(
  age = "25",      # String, not number
  balance = "100"  # String, not number
)

# Correct: Use proper types
alpine_data(
  age = htmlwidgets::JS("25"),      # Number
  balance = htmlwidgets::JS("100")  # Number
)
```

### 3. Validate on Client Only When Appropriate

Use client validation for UX, server validation for security:

```r
# Form with client-side feedback
alpine_tag("form",
  alpine_tag("input",
    type = "email",
    placeholder = "Email"
  ) |>
    alpine_model("email", .lazy = TRUE),
  
  alpine_tag("div",
    "Invalid email" |>
      alpine_show("email && !isValidEmail()")
  ),
  
  alpine_tag("button", "Submit") |>
    htmlwidgets::html_attr("disabled", htmlwidgets::JS("!isValidEmail()"))
) |>
  alpine_data(
    email = "",
    isValidEmail = htmlwidgets::JS("
      function() {
        return /^[^\\s@]+@[^\\s@]+\\.[^\\s@]+$/.test(this.email);
      }
    ")
  )

# ⚠️ ALWAYS validate on server too for security
```

## Debugging & Inspection

### 1. Use Browser Developer Tools

Alpine.js state is accessible in console:

```r
alpine_tag("div") |>
  alpine_data(
    message = "Hello",
    count = 0,
    increment = htmlwidgets::JS("
      function() {
        console.log('Current state:', {
          message: this.message,
          count: this.count
        });
        this.count++;
      }
    ")
  )
```

Open browser console and type: `document.body.__x.$data`

### 2. Log Component State

Add logging to track state changes:

```r
alpine_tag("div",
  alpine_tag("button", "Click me") |>
    alpine_on("click", "logState()")
) |>
  alpine_data(
    clicks = 0,
    logState = htmlwidgets::JS("
      function() {
        this.clicks++;
        console.log('Clicks:', this.clicks);
        console.log('Full state:', this);
      }
    ")
  )
```

### 3. Use Alpine DevTools (Browser Extension)

Install [Alpine.js DevTools](https://chrome.google.com/webstore/detail/alpinejs-devtools) for visual debugging:

- Inspect component state
- Watch state changes in real-time
- Trigger methods from DevTools
- Time travel through state history

## Code Organization

### 1. Keep Components Focused

One responsibility per component:

```r
# Poor: Component doing too many things
alpine_data(
  // Sales data
  sales = list(...),
  // UI state
  selectedTab = "",
  // Filters
  filterCategory = "",
  // Sorting
  sortBy = "",
  // Form data
  formField = "",
  // ... 50 more lines
)

# Better: Component for specific feature
alpine_data(
  sales = list(...),      # One concern
  selectedTab = "recent"
)
```

### 2. Extract Reusable Functions

Create helper functions for complex logic:

```r
# ❌ INLINE: Hard to test and reuse
alpine_data(
  items = list(...),
  processAndFilter = htmlwidgets::JS("
    function() {
      return this.items
        .map(item => ({
          ...item,
          calculated: item.a + item.b
        }))
        .filter(item => item.calculated > 100)
        .sort((a, b) => b.calculated - a.calculated);
    }
  ")
)

# PREFERRED: Reusable functions
alpine_data(
  items = list(...),
  add = htmlwidgets::JS("(a, b) => a + b"),
  processAndFilter = htmlwidgets::JS("
    function() {
      return this.items
        .map(item => ({...item, calculated: this.add(item.a, item.b)}))
        .filter(item => item.calculated > 100)
        .sort((a, b) => b.calculated - a.calculated);
    }
  ")
)
```

### 3. Use Meaningful Names

Clear naming improves readability:

```r
# NOT RECOMMENDED: Vague naming
alpine_data(
  d = list(...),
  fn = htmlwidgets::JS("function() {...}"),
  x = ""
)

# PREFERRED: Clear naming
alpine_data(
  documentList = list(...),
  formatDocument = htmlwidgets::JS("function() {...}"),
  searchQuery = ""
)
```

## Testing Strategies

### 1. Test State Transitions

```r
# Example test pseudo-code
test_that("Counter increments correctly", {
  # Create component with counter
  component <- alpine_tag("div") |>
    alpine_data(count = 0, increment = JS("function() { this.count++; }"))
  
  # Verify initial state
  # Trigger action
  # Verify new state
})
```

### 2. Test Edge Cases

```r
# Test boundary conditions
test_cases <- list(
  list(input = -1, expected = 0),      # Negative
  list(input = 0, expected = 0),       # Zero
  list(input = 1, expected = 1),       # Single
  list(input = 1000000, expected = NA) # Large number
)
```

### 3. Performance Tests

Use browser timing tools:

```r
# Measure operation timing
alpine_data(
  slowOperation = htmlwidgets::JS("
    function() {
      const start = performance.now();
      // ... operation ...
      const duration = performance.now() - start;
      console.log('Operation took ' + duration + 'ms');
    }
  ")
)
```

## Security Best Practices

### 1. Validate All User Input

```r
# Client-side feedback (UX)
alpine_tag("input") |>
  alpine_model("username", .trim = TRUE) |>
  alpine_on("change", "validateUsername()")

# Server-side validation (security)
# Always validate on the server!
```

### 2. Escape Output for XSS Protection

```r
# NOT RECOMMENDED: Renders HTML as-is
alpine_tag("div",
  alpine_attr(`x-bind:innerHTML` = "userInput")
)

# RECOMMENDED: Renders as text
alpine_tag("div",
  alpine_attr(`x-bind:textContent` = "userInput")
)
```

### 3. Use CSP-Compatible Patterns

```r
# NOT RECOMMENDED: Eval with user data
alpine_data(
  expression = JS("eval(userData)")  # Dangerous!
)

# RECOMMENDED: Pre-defined operations
alpine_data(
  operation = "add",  # String, not code
  operations = JS("
    {
      add: (a, b) => a + b,
      subtract: (a, b) => a - b
    }
  ")
)
```

## Common Patterns

### Pattern 1: Loading State

```r
alpine_tag("div",
  alpine_tag("button", "Fetch Data") |>
    alpine_on("click", "fetchData()")
) |>
  alpine_data(
    isLoading = htmlwidgets::JS("false"),
    fetchData = htmlwidgets::JS("
      async function() {
        this.isLoading = true;
        try {
          const response = await fetch('/api/data');
          const data = await response.json();
          // Process data
        } finally {
          this.isLoading = false;
        }
      }
    ")
  )
```

### Pattern 2: Error Handling

```r
alpine_data(
  error = "",
  operation = htmlwidgets::JS("
    async function() {
      this.error = '';
      try {
        // Perform operation
      } catch (e) {
        this.error = e.message;
      }
    }
  ")
)
```

### Pattern 3: Undo/Redo

```r
alpine_data(
  state = htmlwidgets::JS("{ value: 0 }"),
  history = htmlwidgets::JS("[]"),
  undo = htmlwidgets::JS("
    function() {
      if (this.history.length > 0) {
        this.state = this.history.pop();
      }
    }
  ")
)
```

## Checklist: Production Ready

- [ ] All user input validated (client + server)
- [ ] State structure is clear and organized
- [ ] No unnecessary re-renders (debouncing in place)
- [ ] Pagination for large datasets
- [ ] Error handling with user-friendly messages
- [ ] Loading/async states handled
- [ ] Responsive on mobile devices
- [ ] Browser DevTools work without errors
- [ ] CSP-compatible (if using CSP)
- [ ] Performance acceptable on slow connections

## See Also

- [Getting Started](getting-started.qmd)
- [Crosstalk Integration](crosstalk-integration.qmd)
- [Alpine.js Performance Guide](https://alpinejs.dev/)
