---
title: Alpine Plugins: Extending Functionality
---



```{r setup, include=FALSE}
litedown::reactor(echo = TRUE)
RUN_EXAMPLES <- nzchar(Sys.getenv("R_ALPINEJS_RUN_EXAMPLES"))
library(alpinejs)
library(htmltools)
```

```{r test_plugins, eval=RUN_EXAMPLES, echo=FALSE}
# Verify plugin registration works
alpine_setup_persist()
cat("setup_persist() executed successfully\n")
alpine_setup_mask()
cat("setup_mask() executed successfully\n")
```

## Overview

Alpine.js plugins extend the framework with additional directives and utilities. The alpinejs package provides convenient setup functions for the most popular plugins and a system to register custom ones.

**alpinejs provides convenient access to:**

- **persist** -- Save Alpine state to browser localStorage
- **mask** -- Format input fields (phone, credit card, date)
- **sort** -- Add sorting to lists and tables
- **observatory** -- Debug Alpine state in real-time
- Custom plugins from npm or CDN

## Quick Start

### Setup Built-in Plugins

```r
library(alpinejs)
library(htmltools)

# Enable persist plugin
setup_persist()

# Enable mask plugin  
setup_mask()

# Enable sort plugin
setup_sort()

# Check what's registered
get_plugin_registry()
```

```{r test_plugin_registry, eval=RUN_EXAMPLES, echo=FALSE}
# Test that plugin registry functions work
alpine_setup_persist()
alpine_setup_mask()
alpine_setup_sort()

registry <- alpine_get_plugin_registry()
cat("Plugin registry contains:", nrow(registry), "plugins\n")
cat("Registered plugins:", paste(registry$name, collapse = ", "), "\n")
```

### Using Plugins

Once registered, plugins can be used in Alpine components:

```r
# With persist
alpine_setup_persist()

alpine_tag("div",
  alpine_tag("input",
    type = "text",
    placeholder = "Your name",
    alpine_attr(`x-model` = "name")
  ),
  
  "Saved name: ",
  alpine_tag("span",
    alpine_attr(`x-bind:textContent` = "name")
  )
) |>
  alpine_data(
    # With persist, this auto-loads from localStorage
    name = htmlwidgets::JS("localStorage.getItem('name') || ''")
  )
```

## Built-in Plugins

### 1. persist -- localStorage Integration

Automatically save and restore Alpine component state using browser localStorage.

#### Setup

```r
setup_persist()
```

#### Use Cases

- Remember user preferences (theme, layout, language)
- Save form drafts
- Persist filter choices
- Remember sidebar state

#### Example: Remember Theme Preference

```r
library(alpinejs)
library(htmltools)

setup_persist()

alpine_tag("div",
  style = "padding: 20px;",
  
  # Theme toggle
  alpine_tag("button", "Toggle Dark Mode") |>
    alpine_on("click", "darkMode = !darkMode; localStorage.setItem('darkMode', darkMode)"),
  
  # Content
  alpine_tag("div",
    "Welcome! Your theme preference is saved.",
    alpine_attr(`x-bind:style` = "darkMode ? 'background: #222; color: #fff;' : 'background: white'")
  )
) |>
  alpine_data(
    darkMode = htmlwidgets::JS("localStorage.getItem('darkMode') === 'true'")
  ) |>
  htmltools::browsable()
```

#### Data Persistence

localStorage stores data as JSON strings:

```js
// R code
localStorage.setItem('preferences', JSON.stringify({theme: 'dark', lang: 'en'}))

// Later, retrieve:
JSON.parse(localStorage.getItem('preferences'))  // {theme: 'dark', lang: 'en'}
```

### 2. mask -- Input Formatting

Format input fields with patterns (phone, credit card, date, etc.).

#### Setup

```r
setup_mask()
```

#### Common Patterns

| Pattern | Example | Use Case |
|---------|---------|----------|
| `(999) 999-9999` | Phone | US phone numbers |
| `9999-9999-9999-9999` | Credit card | 16-digit cards |
| `99/99/9999` | Date | MM/DD/YYYY format |
| `AAA-999` | Custom | License plates, codes |

#### Example: Phone Number Input

```r
library(alpinejs)
library(htmltools)

setup_mask()

alpine_tag("div",
  style = "padding: 20px;",
  
  alpine_tag("label", "Phone Number:"),
  alpine_tag("input",
    type = "text",
    placeholder = "(123) 456-7890"
  ) |>
    alpine_model("phone") |>
    htmlwidgets::html_attr("x-mask", "(999) 999-9999"),
  
  # Display raw (unformatted) value
  alpine_tag("div",
    "Raw value: ",
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "phone")
    )
  )
) |>
  alpine_data(phone = "") |>
  htmltools::browsable()
```

**How it works:**

- User types: `"1234567890"`
- Mask formats: `"(123) 456-7890"`
- `phone` contains: `"1234567890"` (raw digits)

#### Example: Credit Card Input

```r
alpine_tag("div",
  style = "max-width: 300px; padding: 20px;",
  
  alpine_tag("label", "Card Number:"),
  alpine_tag("input",
    type = "text",
    placeholder = "0000-0000-0000-0000"
  ) |>
    alpine_model("card") |>
    htmlwidgets::html_attr("x-mask", "9999-9999-9999-9999"),
  
  # Display formatted
  alpine_tag("div",
    "Formatted: ",
    alpine_tag("span",
      alpine_attr(`x-bind:textContent` = "cardFormatted()")
    )
  )
) |>
  alpine_data(
    card = "",
    cardFormatted = htmlwidgets::JS("
      function() {
        return this.card.split('-').join(' ');
      }
    ")
  )
```

### 3. sort -- Array Sorting

Add dynamic sorting to lists and tables.

#### Setup

```r
setup_sort()
```

#### Use Cases

- Sortable table headers
- Re-orderable lists
- Dynamic data arrangement
- Multi-column sorting

#### Example: Sortable Table

```r
library(alpinejs)
library(htmltools)

setup_sort()

alpine_tag("table",
  style = "width: 100%; border-collapse: collapse;",
  
  # Headers
  alpine_tag("thead",
    alpine_tag("tr",
      alpine_tag("th", "Name") |>
        alpine_on("click", "sortBy('name')") |>
        htmlwidgets::html_attr("style", "cursor: pointer;"),
      
      alpine_tag("th", "Age") |>
        alpine_on("click", "sortBy('age')") |>
        htmlwidgets::html_attr("style", "cursor: pointer;"),
      
      alpine_tag("th", "City") |>
        alpine_on("click", "sortBy('city')") |>
        htmlwidgets::html_attr("style", "cursor: pointer;")
    )
  ),
  
  # Rows
  alpine_tag("tbody",
    alpine_tag("tr",
      ":for" = "(person, idx) in sortedPeople",
      ":key" = "idx",
      alpine_tag("td",
        alpine_attr(`x-bind:textContent` = "person.name")
      ),
      alpine_tag("td",
        alpine_attr(`x-bind:textContent` = "person.age")
      ),
      alpine_tag("td",
        alpine_attr(`x-bind:textContent` = "person.city")
      )
    )
  )
) |>
  alpine_data(
    people = htmlwidgets::JS("
      [
        {name: 'Alice', age: 28, city: 'NYC'},
        {name: 'Bob', age: 34, city: 'LA'},
        {name: 'Carol', age: 26, city: 'Chicago'}
      ]
    "),
    sortBy = htmlwidgets::JS("
      function(field) {
        this.sortBy = field;
        this.sortedPeople.sort((a, b) => {
          if (a[field] < b[field]) return -1;
          if (a[field] > b[field]) return 1;
          return 0;
        });
      }
    "),
    sortedPeople = htmlwidgets::JS("this.people")
  )
```

## Plugin Manager Functions

### `setup_persist()`

Register the persist plugin for localStorage support:

```r
setup_persist()

# Verify registration
is_plugin_registered("persist")  # TRUE
```

### `setup_mask()`

Register mask plugin for input formatting:

```r
setup_mask()

is_plugin_registered("mask")  # TRUE
```

### `setup_sort()`

Register sort plugin for array sorting:

```r
setup_sort()

is_plugin_registered("sort")  # TRUE
```

### `register_alpine_plugin()`

Register a custom plugin from npm or CDN:

```r
# From npm (requires @-scope)
register_alpine_plugin(
  name = "my-custom",
  version = "1.0.0",
  npm_scope = "my-org"  # Becomes @my-org/my-custom
)

# From CDN (advanced)
register_alpine_plugin(
  name = "from-cdn",
  version = "2.0.0",
  src = "https://cdn.example.com/plugin.js"
)
```

### `is_plugin_registered()`

Check if a plugin is registered:

```r
is_plugin_registered("persist")  # TRUE/FALSE
is_plugin_registered("mask")     # TRUE/FALSE
```

### `get_plugin_registry()`

View all registered plugins:

```r
registry <- get_plugin_registry()
str(registry)

# Output example:
# List of 3
#  $ persist: list(version = "1.0.0", ...)
#  $ mask   : list(version = "1.0.0", ...)
#  $ sort   : list(version = "1.0.0", ...)
```

## Advanced: Custom Plugins

### Registering from npm

If you have a custom Alpine.js plugin on npm:

```r
# My company publishes @acme/analytics on npm
register_alpine_plugin(
  name = "analytics",
  version = "1.2.3",
  npm_scope = "acme"  # Becomes @acme/analytics
)
```

### CSP Considerations

Plugins must be CSP-compatible (no `eval` or inline scripts):

```r
# Tell alpinejs about CSP compliance
register_alpine_plugin(
  name = "safe-plugin",
  version = "1.0.0",
  npm_scope = "org",
  csp = TRUE  # CSP-compatible
)
```

## Real-World Examples

### Example 1: Todo List with Persistence

```r
library(alpinejs)
library(htmltools)

alpine_setup_persist()

alpine_tag("div",
  style = "max-width: 500px; padding: 20px;",
  
  alpine_tag("h2", "Todo List (Auto-saved)"),
  
  # Input
  alpine_tag("div",
    style = "display: flex; gap: 10px; margin-bottom: 20px;",
    alpine_tag("input",
      type = "text",
      placeholder = "New task...",
      alpine_attr(`x-model` = "newTask")
    ),
    
    alpine_tag("button", "Add",
      alpine_attr(`@click` = "addTask()")
    )
  ),
  
  # List
  alpine_tag("ul",
    alpine_tag("li",
      ":for" = "(task, idx) in tasks",
      ":key" = "idx",
      alpine_tag("span",
        alpine_attr(`x-bind:textContent` = "task")
      ),
      alpine_tag("button", "Delete",
        alpine_attr(`@click` = "removeTask(idx)")
      )
    )
  )
) |>
  alpine_data(
    tasks = htmlwidgets::JS("
      JSON.parse(localStorage.getItem('todos')) || []
    "),
    newTask = "",
    addTask = htmlwidgets::JS("
      function() {
        if (!this.newTask.trim()) return;
        this.tasks.push(this.newTask);
        localStorage.setItem('todos', JSON.stringify(this.tasks));
        this.newTask = '';
      }
    "),
    removeTask = htmlwidgets::JS("
      function(idx) {
        this.tasks.splice(idx, 1);
        localStorage.setItem('todos', JSON.stringify(this.tasks));
      }
    ")
  ) |>
  htmltools::browsable()
```

### Example 2: Formatted Payment Form

```r
library(alpinejs)
library(htmltools)

alpine_setup_mask()

alpine_tag("form",
  style = "max-width: 400px; padding: 20px;",
  
  # Card number
  alpine_tag("div",
    alpine_tag("label", "Card Number:"),
    alpine_tag("input",
      type = "text",
      placeholder = "0000-0000-0000-0000",
      alpine_attr(
        `x-model` = "card",
        `x-mask` = "9999-9999-9999-9999"
      )
    )
  ),
  
  # Expiry
  alpine_tag("div",
    alpine_tag("label", "Expires:"),
    alpine_tag("input",
      type = "text",
      placeholder = "MM/YY",
      alpine_attr(
        `x-model` = "expiry",
        `x-mask` = "99/99"
      )
    )
  ),
  
  # CVC
  alpine_tag("div",
    alpine_tag("label", "CVC:"),
    alpine_tag("input",
      type = "text",
      placeholder = "000",
      alpine_attr(
        `x-model` = "cvc",
        `x-mask` = "999"
      )
    )
  ),
  
  button("Pay") |>
    alpine_on("click", "processPayment()")
) |>
  alpine_data(
    card = "",
    expiry = "",
    cvc = "",
    processPayment = htmlwidgets::JS("
      function() {
        console.log({card: this.card, expiry: this.expiry, cvc: this.cvc});
      }
    ")
  )
```

### Example 3: Preferences with Persistence

```r
library(alpinejs)
library(htmltools)

setup_persist()

alpine_tag("div",
  style = "padding: 20px;",
  
  alpine_tag("h3", "User Preferences"),
  
  # Theme
  div(
    label("Theme:"),
    select(
      option(value = "light", "Light"),
      option(value = "dark", "Dark"),
      option(value = "auto", "Auto")
    ) |>
      alpine_model("theme") |>
      alpine_on("change", "savePreferences()")
  ),
  
  # Language
  div(
    label("Language:"),
    select(
      option(value = "en", "English"),
      option(value = "es", "Español"),
      option(value = "fr", "Français")
    ) |>
      alpine_model("language") |>
      alpine_on("change", "savePreferences()")
  ),
  
  # Notifications
  div(
    label("Notifications:"),
    input(type = "checkbox") |>
      alpine_model("notifications") |>
      alpine_on("change", "savePreferences()")
  ),
  
  div("Preferences saved to browser!")
) |>
  alpine_data(
    theme = htmlwidgets::JS("
      localStorage.getItem('pref_theme') || 'light'
    "),
    language = htmlwidgets::JS("
      localStorage.getItem('pref_language') || 'en'
    "),
    notifications = htmlwidgets::JS("
      localStorage.getItem('pref_notifications') === 'true'
    "),
    savePreferences = htmlwidgets::JS("
      function() {
        localStorage.setItem('pref_theme', this.theme);
        localStorage.setItem('pref_language', this.language);
        localStorage.setItem('pref_notifications', this.notifications);
      }
    ")
  )
```

## Plugin Performance

Plugins are loaded once per page:

- No performance penalty for multiple components
- Lazy-loaded when first component using plugin renders
- Minimal bundle size impact

## Troubleshooting

**Q: Plugin not working even after setup?**
A: Verify with `is_plugin_registered()`. Reload page if freshly added.

**Q: Mask input too restrictive?**
A: Use looser patterns. `x-mask` is optional; users can enter any characters.

**Q: localStorage full errors?**
A: Clear old data or reduce what you're persisting.

**Q: Plugin breaks in production?**
A: Check CSP settings. Use CSP-compatible plugins (persist, mask are safe).

## See Also

- [Alpine.js Official Plugins](https://alpinejs.dev/)
- [Getting Started](getting-started.qmd)
- [Input Modifiers Guide](input-modifiers.qmd)
