---
title: Reactivity and Global State
---

```{r setup, include=FALSE}
litedown::reactor(echo = TRUE)
RUN_EXAMPLES <- nzchar(Sys.getenv("R_ALPINEJS_RUN_EXAMPLES"))
library(alpinejs)
library(htmltools)
```

```{r test_magics_globals, eval=TRUE, echo=FALSE}
# Verify new functions are available
fns <- c("alpine_watch", "alpine_dispatch", "alpine_next_tick",
         "alpine_refs", "alpine_store_js", "alpine_el",
         "alpine_global_data", "alpine_global_store",
         "alpine_effect", "alpine_ref", "alpine_if",
         "alpine_cloak", "alpine_ignore", "alpine_modelable",
         "alpine_transition", "alpine_teleport", "alpine_id")
missing <- fns[!vapply(fns, existsMethod, logical(1), signature = character(0))]
missing <- fns[!vapply(fns, function(f) exists(f, mode = "function"), logical(1))]
cat("All new functions available:", length(missing) == 0, "\n")
```

# Reactivity and Global State

This vignette covers Alpine.js magic properties and global methods -- the tools
for advanced reactivity: watching data, dispatching events between components,
and sharing state across a page.

- **Magic properties** (`$watch`, `$dispatch`, `$nextTick`, `$refs`, `$store`, `$el`)
  are available inside any Alpine expression.
- **Global methods** (`Alpine.data()`, `Alpine.store()`) register reusable components
  and cross-component shared state before Alpine initializes.

---

## Magic Properties

Magic properties are special variables prefixed with `$` that are available inside
Alpine.js expressions. The alpinejs package provides helper functions that generate
the corresponding JavaScript expression strings.

These helpers return `htmlwidgets::JS()` objects, so they can be used anywhere an
expression string is accepted: in `alpine_init()`, `alpine_on()`, `alpine_bind()`,
and other directives.

### $watch -- React to Property Changes

`alpine_watch("property", callback)` generates a `$watch('property', callback)`
expression that runs the callback whenever the named property changes. Use it
inside `alpine_init()` to set up watchers when the component first loads.

```{r}
# Log every change to the search query
tagList(
  tags$div(
    tags$input(type = "text", placeholder = "Search...") |>
      alpine_model("query"),
    tags$p() |>
      alpine_text("'Searching for: ' + query")
  ) |>
    alpine_init(alpine_watch("query", "value => console.log('Query:', value)")) |>
    alpine_data(query = "")
)
```

You can watch deeply nested properties using dot notation, and receive the old
value as the second callback argument:

```{r}
# Watch a nested property (console only -- no visible output)
tags$div() |>
  alpine_init(
    alpine_watch("user.name", "(newVal, oldVal) => console.log(newVal, oldVal)")
  ) |>
  alpine_data_js("{ user: { name: 'Alice' } }")
```

### $dispatch -- Send Events Between Components

`alpine_dispatch("event-name", detail)` generates a `$dispatch('event', detail)`
expression that emits a custom DOM event. Other elements can listen for it with
`alpine_on("event-name.window", ...)` (the `.window` modifier is needed for
cross-component communication since events bubble to the window).

```{r}
# Component 1: dispatch an event when the button is clicked
emitter <- tags$div(
  style = "padding: 10px; border: 1px solid #ccc;",
  tags$p("Component 1 -- Emitter"),
  tags$button("Notify Other Component") |>
    alpine_on("click", alpine_dispatch("notify", list(message = "Hello!")))
) |>
  alpine_data(x = 0)

# Component 2: listen for the event from Component 1
listener <- tags$div(
  style = "padding: 10px; border: 1px solid #ccc; margin-top: 8px;",
  tags$p("Component 2 -- Listener"),
  tags$p() |>
    alpine_text("lastMessage || 'No messages yet'")
) |>
  alpine_init(
    "$nextTick(() => {})"  # placeholder init
  ) |>
  alpine_on("notify.window", "lastMessage = $event.detail.message") |>
  alpine_data(lastMessage = "")

tagList(emitter, listener)
```

### $nextTick -- Act After DOM Updates

`alpine_next_tick(callback)` generates a `$nextTick(callback)` expression that
defers the callback until after Alpine has finished processing the current reactive
update and updated the DOM. Use it when you need to measure or interact with the
DOM after a state change.

```{r}
# Focus an input after toggling it visible
tags$div(
  tags$button("Show and focus input") |>
    alpine_on("click", paste0(
      "show = true; ",
      alpine_next_tick("() => $refs.myInput.focus()")
    )),
  tags$input(type = "text", placeholder = "Now focused") |>
    alpine_ref("myInput") |>
    alpine_show("show")
) |>
  alpine_data(show = FALSE)
```

### $refs -- Access DOM Elements

`alpine_refs("name")` generates a `$refs.name` expression for the element labeled
with `alpine_ref()`. Use it to read values, call DOM methods, or pass elements
to other functions.

```{r}
# Copy input value to clipboard
tags$div(
  tags$input(type = "text", value = "Copy me!") |>
    alpine_ref("copyTarget"),
  tags$button("Copy to Clipboard") |>
    alpine_on("click", paste0(
      "navigator.clipboard.writeText(",
      alpine_refs("copyTarget"),
      ".value)"
    )),
  tags$span() |>
    alpine_text("copied ? 'Copied!' : ''")
) |>
  alpine_on("click", "copied = true; setTimeout(() => copied = false, 2000)",
            .cssSelector = "button") |>
  alpine_data(copied = FALSE)
```

You can also just write `$refs.name` inline -- the helper is provided for
programmatic expression building when the ref name is an R variable.

### $store -- Access Global Stores

`alpine_store_js("store_name", "property")` generates a `$store.name.property`
expression for reading from or modifying a global Alpine store. See the
[Global Methods](#global-methods) section below for registering stores.

```{r, eval=FALSE}
# Read from a store
alpine_store_js("theme")            # -> $store.theme
alpine_store_js("theme", "mode")    # -> $store.theme.mode
```

### $el -- Reference the Root Element

`alpine_el()` generates `$el`, which refers to the root DOM element of the
current Alpine component. Useful for passing the element to third-party libraries
or reading its dimensions.

```{r, eval=FALSE}
# Scroll the component into view
tags$button("Scroll here") |>
  alpine_on("click", paste0(alpine_el(), ".scrollIntoView({ behavior: 'smooth' })"))
```

---

## x-effect: Reactive Side Effects

`alpine_effect()` wraps an expression that Alpine.js re-evaluates automatically
whenever any reactive data it references changes. Unlike `$watch`, you do not
specify which property to watch -- Alpine tracks all reactive data accessed.

```{r}
# Log query to console on every change (side effect)
tags$div(
  tags$input(type = "text", placeholder = "Type to search...") |>
    alpine_model("query"),
  tags$p(style = "color: gray; font-size: 0.9em;",
    "Open browser console to see reactive logging."
  )
) |>
  alpine_effect("console.log('[effect] query is now:', query)") |>
  alpine_data(query = "")
```

---

## Global Methods

### Alpine.data() -- Reusable Components

`alpine_global_data(name, js_definition)` registers a named, reusable Alpine
component that can be used by name in any `x-data` attribute. This avoids
repeating the same component definition when using it multiple times on a page.

```{r}
# Register a reusable counter component
counter_registration <- alpine_global_data(
  "counter",
  "() => ({
    count: 0,
    increment() { this.count++ },
    decrement() { if (this.count > 0) this.count-- },
    reset()     { this.count = 0 }
  })"
)

# Use it twice -- each instance has its own state
counter1 <- tags$div(
  style = "display: flex; gap: 8px; align-items: center;",
  tags$span("Counter A:"),
  tags$button("-") |> alpine_on("click", "decrement()"),
  tags$span() |> alpine_text("count"),
  tags$button("+") |> alpine_on("click", "increment()"),
  tags$button("Reset") |> alpine_on("click", "reset()")
) |> alpine_data_js("counter")

counter2 <- tags$div(
  style = "display: flex; gap: 8px; align-items: center; margin-top: 8px;",
  tags$span("Counter B:"),
  tags$button("-") |> alpine_on("click", "decrement()"),
  tags$span() |> alpine_text("count"),
  tags$button("+") |> alpine_on("click", "increment()"),
  tags$button("Reset") |> alpine_on("click", "reset()")
) |> alpine_data_js("counter")

tagList(counter_registration, counter1, counter2)
```

Each counter has completely independent state -- changing counter A does not
affect counter B.

### Alpine.store() -- Global Shared State

`alpine_global_store(name, ...)` registers a global store accessible from any
Alpine component on the page via `$store.name`. When any component writes to the
store, all other components that read from it update reactively.

#### Example: Shared Theme Toggle

```{r}
# Register a dark mode store with a toggle method (raw JS)
theme_store <- alpine_global_store(
  "theme",
  htmlwidgets::JS("{ dark: false, toggle() { this.dark = !this.dark } }")
)

# Component A: the toggle button
toggle_btn <- tags$div(
  tags$button() |>
    alpine_text("$store.theme.dark ? 'Switch to Light' : 'Switch to Dark'") |>
    alpine_on("click", "$store.theme.toggle()") |>
    alpine_bind("style", "$store.theme.dark ? 'background:#222;color:#fff' : ''")
) |>
  alpine_data_js("{}")

# Component B: themed content panel (completely separate component)
themed_panel <- tags$div(
  tags$p() |>
    alpine_text("'Current mode: ' + ($store.theme.dark ? 'Dark' : 'Light')"),
  style = "padding: 10px; margin-top: 8px; border: 1px solid #ccc;"
) |>
  alpine_bind("style", "$store.theme.dark ? 'background:#1a1a1a;color:#fff' : ''") |>
  alpine_data_js("{}")

tagList(theme_store, toggle_btn, themed_panel)
```

Clicking the toggle button in Component A instantly updates Component B, even
though they share no parent `x-data` component -- the store is the shared channel.

#### Example: Notification Queue (R data)

Stores can also be initialized from R data:

```{r}
notif_store <- alpine_global_store(
  "notifications",
  count   = 0L,
  visible = FALSE
)

notif_ui <- tags$div(
  tags$button("Trigger Notification") |>
    alpine_on("click",
      "$store.notifications.count++; $store.notifications.visible = true;
       setTimeout(() => $store.notifications.visible = false, 3000)"
    ),
  tags$div(
    style = "background: #28a745; color: white; padding: 8px 12px; 
             margin-top: 8px; border-radius: 4px;",
    "Notification sent! Total: ",
    tags$span() |> alpine_text("$store.notifications.count")
  ) |>
    alpine_show("$store.notifications.visible") |>
    alpine_transition(.opacity = TRUE, .duration = 300)
) |>
  alpine_data_js("{}")

tagList(notif_store, notif_ui)
```

---

## Complete Example: Search with Effects and Refs

This example combines several features: `$watch` via `x-effect`, `x-ref`, and
`alpine_if` for conditional rendering of results.

```{r}
search_demo <- tags$div(
  style = "font-family: sans-serif; max-width: 400px;",

  # Search input with ref for auto-focus
  tags$input(
    type = "text",
    placeholder = "Search items...",
    style = "width: 100%; padding: 6px; box-sizing: border-box;"
  ) |>
    alpine_model("query", .debounce = 200) |>
    alpine_ref("searchInput"),

  # Count display
  tags$p(style = "color: gray; font-size: 0.85em;") |>
    alpine_text("filteredItems.length + ' results'"),

  # Results list
  alpine_for(
    tags$div(
      style = "padding: 4px 0; border-bottom: 1px solid #eee;",
      tags$span() |> alpine_text("item")
    ),
    "item in filteredItems",
    key = "item"
  ),

  # Empty state
  alpine_if(
    tags$p(style = "color: #999;", "No items matched."),
    "filteredItems.length === 0 && query.length > 0"
  )
) |>
  alpine_init(alpine_next_tick("() => $refs.searchInput.focus()")) |>
  alpine_effect("console.log('[effect] Visible results:', filteredItems.length)") |>
  alpine_data_js("{
    query: '',
    items: ['apples', 'bananas', 'cherries', 'dates', 'elderberries', 'figs'],
    get filteredItems() {
      if (!this.query) return this.items;
      return this.items.filter(i => i.startsWith(this.query.toLowerCase()));
    }
  }")

search_demo
```

---

## Reference: Magic Property Helpers

| Function                        | Generated Expression     | Purpose                               |
|:--------------------------------|:-------------------------|:--------------------------------------|
| `alpine_watch("p", "cb")`       | `$watch('p', cb)`        | Watch a property for changes          |
| `alpine_dispatch("ev", detail)` | `$dispatch('ev', {...})` | Dispatch a custom DOM event           |
| `alpine_next_tick("cb")`        | `$nextTick(cb)`          | Run code after DOM update             |
| `alpine_refs("name")`           | `$refs.name`             | Access a labeled DOM element          |
| `alpine_store_js("n", "p")`    | `$store.n.p`             | Read/write a global store property    |
| `alpine_el()`                   | `$el`                    | Access the component root element     |

## Reference: Global Method Functions

| Function                             | Alpine.js Equivalent   | Purpose                             |
|:-------------------------------------|:-----------------------|:------------------------------------|
| `alpine_global_data("name", jsDef)` | `Alpine.data('name', factory)` | Register a reusable component |
| `alpine_global_store("name", ...)`  | `Alpine.store('name', data)`  | Register global shared state  |

Both functions return `htmltools::tagList` containing a `<script>` tag and the
Alpine.js dependency. Include the result anywhere in your page HTML.

---

## Next Steps

- **Input Modifiers**: See the [Input Modifiers Vignette](input-modifiers.qmd) for `.lazy`, `.debounce`, `.number`, and `.trim`
- **Plugin Ecosystem**: See the [Plugin Ecosystem Vignette](plugin-ecosystem.qmd) for Persist, Mask, and Sort plugins
- **Crosstalk Integration**: See the [Crosstalk Integration Vignette](crosstalk-integration.qmd) for cross-widget reactivity
